---
title: "Michelson Speed-of-light Measurements"
author: "Lucien Lin"
date: 2020-
output: 
  github_document:
    toc: true
prerequisites:
  - e-data02-derive
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: When studying physical problems, there is an important
distinction between *error* and *uncertainty*. The primary purpose of
this challenge is to dip our toes into these factors by analyzing a real
dataset.

*Reading*: [Experimental Determination of the Velocity of
Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115)
(Optional)

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and
supporting files (`report_files/` folder) when you are done! Then submit
a link to Canvas. **Your Challenge submission is not complete without
all files uploaded to GitHub.**

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM    <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00  # Michelson's speed estimate (km / s)
LIGHTSPEED_PM        <- 51         # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to
measure the speed of light. His approach was a development upon the
method of Foucault[3], and resulted in a new estimate of
$v_0 = 299944 \pm 51$ kilometers per second (in a vacuum). This is very
close to the modern *exact* value of `r LIGHTSPEED_VACUUM`. In this
challenge, you will analyze Michelson's original data, and explore some
of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code
chunk below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the
value `LIGHTSPEED_VACUUM`) because the meter is actually
[*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition)
in terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse()
```

*Data dictionary*:

-   `Date`: Date of measurement
-   `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair,
    1 = poor
-   `Temp`: Ambient temperature (Fahrenheit)
-   `Velocity`: Measured speed of light (km / s)

### **q1** Re-create the following table (from Michelson (1880), pg. 139) using `df_michelson` and `dplyr`. Note that your values *will not* match those of Michelson *exactly*; why might this be?

| Distinctness | n   | MeanVelocity |
|--------------|-----|--------------|
| 3            | 46  | 299860       |
| 2            | 39  | 299860       |
| 1            | 15  | 299810       |

```{r q1-task}
## TODO: Compute summaries
df_q1 <- 
  df_michelson %>%
    group_by(Distinctness) %>%
    summarize(
      n = n(),
      MeanVelocity = mean(Velocity)
    )
df_q1 %>%
  arrange(desc(Distinctness)) %>%
  knitr::kable()
```

**Observations**: - Write your observations here! - My velocity values
contain decimals, but I was able to group the data by distinctness and
count the number of rows of each group. - Why might your table differ
from Michelson's? - Michelson may have rounded his values, or chosen his
precision not to include decimals.

The `Velocity` values in the dataset are the speed of light *in air*;
Michelson introduced a couple of adjustments to estimate the speed of
light in a vacuum. In total, he added $+92$ km/s to his mean estimate
for `VelocityVacuum` (from Michelson (1880), pg. 141). While the
following isn't fully rigorous ($+92$ km/s is based on the mean
temperature), we'll simply apply this correction to all the observations
in the dataset.

### **q2** Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to `Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
## TODO: Adjust the data, assign to df_q2
df_q2 <- 
  df_michelson %>%
    mutate(
      VelocityVacuum = Velocity + 92
    )

df_q2
```

As part of his study, Michelson assessed the various potential sources
of error, and provided his best-guess for the error in his
speed-of-light estimate. These values are provided in
`LIGHTSPEED_MICHELSON`---his nominal estimate---and
`LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently,
Michelson believed the true value of the speed-of-light probably lay
between `LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and
`LIGHTSPEED_MICHELSON + LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

-   **Error** is the difference between a true value and an estimate of
    that value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
-   **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does
not know the error. The best they can do is quantify their degree of
uncertainty. We will learn some means of quantifying uncertainty in this
class, but for many real problems uncertainty includes some amount of
human judgment.[2]

### **q3** Compare Michelson's speed of light estimate against the modern speed of light value. Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?

```{r q3-task}
## TODO: Compare Michelson's estimate and error against the true value
## Your code here!

LIGHTSPEED_MICHELSON - LIGHTSPEED_VACUUM

LIGHTSPEED_MICHELSON - LIGHTSPEED_PM
LIGHTSPEED_MICHELSON + LIGHTSPEED_PM
LIGHTSPEED_MICHELSON 

LIGHTSPEED_VACUUM
LIGHTSPEED_PM * 2
```

**Observations**: - Is Michelson's estimate of the error (his
uncertainty) greater or less than the true error? - Michelson's error
estimate is less than the true error. - Make a quantitative comparison
between Michelson's uncertainty and his error. - The true error was
151.542 km/s. Michelson's uncertainty causes him to give an error
estimate of +/- 56, with a full range of 102 km/s. This mean that
Michelson's uncertainty wasn't able to capture the true value of the
uncertainty.

The following plot shows all of Michelson's data as a [control
chart](https://en.wikipedia.org/wiki/Control_chart); this sort of plot
is common in manufacturing, where it is used to help determine if a
manufacturing process is under [statistical
control](https://en.wikipedia.org/wiki/Statistical_process_control).
Each dot is one of Michelson's measurements, and the grey line connects
the mean taken for each day. The same plot also shows simulated data
using a probability model. We'll get into statistics later in the
course; for now, let's focus on understanding what real and simulated
data tend to look like.

### **q4** Inspect the following plot with the `Real` Michelson data and `Simulated` data from a probability model. Document the similarities and differences between the data under *observe* below.

```{r q4-cf-real-simulated}
## Note: No need to edit this chunk!
## Calibrate simulated data
v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Simulated, Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%

  ggplot(aes(Date, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +

  geom_line(
    data = . %>%
      group_by(Date, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Date of Measurement (1879)",
    y = "Velocity (in Vacuum)"
  )
```

**Observations**: Similarities - The general trend is replicated
similarly in the simulated model, where the mean of greater measurements
were measured on similar dates. Overall, both simulated and real control
graphs were shown to be within the bounds set by Michelson's error
estimate. Differences - I noticed that in the real data, there were more
outliers that went outside of the error estimates, which leads me to
wonder why the error estimate wasn't bigger to account for the outliers.
The simulated model had means that were mostly contained within the
error measurements.

### **q5** You have access to a few other variables. Construct a **at least three** visualizations of `VelocityVacuum` against these other factors. Are there other patterns in the data that might help explain the difference between Michelson's estimate and `LIGHTSPEED_VACUUM`?

```{r}
df_q2 %>%
  ggplot(
    aes(
      x = Temp,
      y = VelocityVacuum,
      color = Distinctness
    )
  ) + 
  geom_point() +
  geom_line(
    data = . %>%
      group_by(Temp) %>%
      summarize(velocity_mean = mean(VelocityVacuum)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) + 
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  )

df_q2 %>%
  ggplot(
    aes(
      x = Distinctness,
      y = VelocityVacuum,
      color = Temp
    )
  ) +
  geom_point()

df_q2 %>%
  filter(Distinctness == "3") %>%
  ggplot(
    aes(
      x = Temp,
      y = VelocityVacuum
    )
  ) +
  geom_point() +
  geom_line(
    data = . %>%
      group_by(Temp) %>%
      summarize(velocity_mean = mean(VelocityVacuum)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) + 
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  )

df_q2 %>%
  filter(Distinctness == "3") %>%
  ggplot(
    aes(
      x = Date,
      y = VelocityVacuum
    )
  ) +
  geom_point() +
  geom_line(
    data = . %>%
      group_by(Date) %>%
      summarize(velocity_mean = mean(VelocityVacuum)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) + 
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  )
```

**Observations**:

-   Make sure to record observations on your graphs!
-   Graph #1 plots VelocityVacuum against the Temp. The mean velocity at
    each temperature measurement is also plotted, with each point being
    specified by its distinctness. When using temperature to compare the
    velocities in a vacuum, we can see that the overall mean trend of
    the velocities are mainly contained within the errors, with some
    outliers that give lower velocity values. Since the velocities of
    the vacuum were extracted from the velocities measured with air, I
    thought that temperature may have an effect on the velocity. When
    comparing this graph to that of the time, I noticed that the values
    are very similar, though the overall maximum velocity is lower when
    measuring temperature.
-   Graph #2 plots VelocityVacuum against the Distinctness. While there
    is severe overlap, I noticed that there were more measurements for
    Distinctness 2 and 3. Depending on the distinctness during the
    measurement phase, the measured velocities may be more
    accurate/inaccurate.
-   Graph #3 is similar to Graph #1, but I have filtered out the points
    to only have a Distinctness of 3. Due to the lack of data points,
    the mean trendline oscillates between high and low mean velocities.
    I notice here that for temperatures around 75F, the velocities are
    lower than that specified by the error bounds.
-   Graph #4 is similar to the graph for question 4, but I have also
    filtered out the distinctness to only show the distinctness of 3.
    The lack of data points causes the trend line to fluctuate quite a
    bit, but I notice that the earlier dates yielded higher velocities
    while the later dates yielded lower velocities.
-   Comparing Michelson's estimate against the value in the vacuum, I'm
    noticing that the time and temperature may directly affect how
    accurate the measured velocities are. I think it would be
    interesting to find a correlation between time and temperature, as I
    don't think Michelson was able to maintain the same temperature when
    taking measurements.

## Bibliography

-   [1] Michelson, [Experimental Determination of the Velocity of
    Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115)
    (1880) 
-   [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical
    Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf)
    (1986) 
-   [3] BYU video about a [Fizeau-Foucault
    apparatus](https://www.youtube.com/watch?v=Ik5ORaaeaME), similar to
    what Michelson used.
