---
title: "Gapminder"
author: "Lucien Lin"
date: 3/2/2025
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)

names(gapminder)
```

**Observations**:

-   Country
-   Continent
-   Year
-   LifeExp
-   Pop
-   gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- 
    gapminder %>%
      pull(year) %>%
      max()

year_max

year_min <- 
    gapminder %>%
      pull(year) %>%
      min()

year_min
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
df_q2_a <- 
gapminder %>%
  filter(
    year == year_min
  ) %>% 
  ggplot(
    aes(
      x = continent, 
      y = gdpPercap
    )
  ) + 
  geom_boxplot() 

df_q2_b <-
  gapminder %>%
  filter(
    year == year_min, 
    gdpPercap < 15000
  ) %>% 
  ggplot(
    aes(
      x = continent, 
      y = gdpPercap
    )
  ) + 
  geom_boxplot() 

df_q2_a
df_q2_b
```

**Observations**:

-   Write your observations here
-   Asia has a massive outlier that causes the data to be squished
-   Europe has the greatest distribution, while Oceania has the lowest distribution
-   The lowest medium is Africa

**Difficulties & Approaches**:

-   Write your challenges and your approach to solving them
-   When I generated the first plot, I had trouble reading the graph because everything compressed to allow for the Asia outlier to be displayed. To account for this, I generated a new plot that filtered out that outlier to get a better look at the boxplot data. In this case, I think it was okay to do so since I wanted to analyze spread and the averages, rather than discussing trends in the outliers.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder %>%
  filter(year == year_min) %>%
  group_by(continent) %>%
  mutate(
    q1 = quantile(gdpPercap, 0.25),
    q3 = quantile(gdpPercap, 0.75),
    IQR = q3 - q1, 
    lower_bound = q1 - 1.5 * IQR, 
    upper_bound = q3 + 1.5 * IQR
  ) %>%
  ungroup() %>%
  filter(
    lower_bound > gdpPercap | upper_bound < gdpPercap
  ) %>%
  select(
    country, continent, gdpPercap, everything()
  ) %>%
  arrange(
    desc(gdpPercap)
  ) 
```

**Observations**:

-   Identify the outlier countries from q2
    -   (Your response here)
    -   Kuwait
    -   Switzerland
    -   United States

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(
    year %in% c(year_min, year_max)
  ) %>%
  ggplot(
    aes(
      x = continent, 
      y = gdpPercap,
      fill = as.factor(year)
    )
  ) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("Kuwait", "Switzerland", "United States")),
    mapping = aes(color = country),
    position = position_dodge(width = 0.8), 
    size = 2
  )
```

**Observations**:

-   Write your observations here
-   Generatlly, we see that the average GDP across all the contents have increases in 2007. Country-wise, this graph shows that the United States and Switzerland have their GDP increased, but Kuwait had their GDP dropped. The 2007 boxplots also display a greater distribtion when compared to the 1952 variants. Since I don't know what was going on during those years, I would prefer to explain the distribution differences to the data variety present for both years. I see that the outliers of each country seem to remain outliers in both years, except for Switzerland.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder %>%
  filter(
    year %in% c(year_min, year_max)
  ) %>%
  ggplot(
    aes(
      x = continent, 
      y = lifeExp,
      fill = as.factor(year)
    )
  ) +
  geom_boxplot()
```

-   (Your notes and observations here)
-   There is a positive trend with life expectancy across all the continents. The data distribution for Oceania is extremely low. This may be due to lack of data to display. Asia seems to have an even distribution, while the Americas and Europe had tigher distributions during 2007.

```{r q5-task2}
## TASK: Your second graph
gapminder %>%
  mutate(year = as.factor(year)) %>%
  filter(
    year %in% c(1952, 1962, 1972, 1982, 1992, 2007)
  ) %>%
  ggplot(
    aes(
      x = continent, 
      y = gdpPercap,
      fill = as.factor(year)
    )
  ) +
  geom_boxplot()
  
```

-   (Your notes and observations here)
-   This boxplot looks at the trends of how each continents' GDP per Capita increases every \~10 years. I see that generally, the trend is increasing, but developing a linear regression model for the data might be difficult since certain continents such as Asia and Europe have an increased distribution compared to the others. Unless I outline what country the specific outliers are, I can't determine whether the countries themselves have increased or decreased.

```{r q5-task3}
## TASK: Your third graph
gapminder %>%
  filter(
    country == "Kuwait"
  ) %>%
  ggplot(
    aes(
      x = year,
      y = gdpPercap
    )
  ) + 
  geom_line()

```

-   This is a line plot that I wanted to look at the trends of the outliers that I identified in a previous problem. I noticed that while all the continents demonstrated a positive growth in GDP per Capita, Kuwait was a country that demonstrated negative growth. Perhaps something that I would like to look further for this plot is to find the reasoning behind this GDP per Capita decrease. I speculate that this may be due to population, which is a graph that I would like to look at.
