---
title: "Antibiotics"
author: "Lucien Lin"
date: 3/9/2025
output:
  github_document
editor_options: 
  markdown: 
    wrap: sentence
---

*Purpose*: Creating effective data visualizations is an *iterative* process; very rarely will the first graph you make be the most effective.
The most effective thing you can do to be successful in this iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely dependent on *the question you are trying to answer*.
A visual that is totally ineffective for one question may be perfect for answering a different question.

In this challenge, you will practice *iterating* on data visualization, and will anchor the *assessment* of your visuals using two different questions.

*Note*: Please complete your initial visual design **alone**.
Work on both of your graphs alone, and save a version to your repo *before* coming together with your team.
This way you can all bring a diversity of ideas to the table!

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**.
The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done!
Then submit a link to Canvas.
**Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(ggrepel)
```

*Background*: The data[1] we study in this challenge report the [*minimum inhibitory concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration) (MIC) of three drugs for different bacteria.
The smaller the MIC for a given drug and bacteria pair, the more practical the drug is for treating that particular bacteria.
An MIC value of *at most* 0.1 is considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin, streptomycin, and neomycin---on 16 bacteria.
Bacteria are categorized into a genus based on a number of features, including their resistance to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```

# Visualization

<!-- -------------------------------------------------- -->

### **q1** Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the data** `df_antibiotics`.
These **cannot** be simple variations on the same graph; for instance, if two of your visuals could be made identical by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16 bacteria*.
For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of ideas, and don't fixate on any one idea for too long.
You will have a chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be helpful to `pivot_longer()` the data to make certain visuals easier to construct.

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*.
This means **it must be possible to identify each of the 16 bacteria by name.** You must also show whether or not each bacterium is Gram positive or negative.

```{r q1.1}
# WRITE YOUR CODE HERE
df_antibiotics %>%
  pivot_longer(
    names_to = "drugs",
    values_to = "concentrations",
    cols = c(penicillin, streptomycin, neomycin)
  ) %>%
  ggplot(
    aes(
      x = concentrations,
      y = bacteria,
      color = drugs,
      shape = gram
    )
  )+
  scale_x_log10() + 
  geom_point()

```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*.
This means **it must be possible to identify each of the 16 bacteria by name.** You must also show whether or not each bacterium is Gram positive or negative.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.2}
# WRITE YOUR CODE HERE
df_antibiotics %>%
  pivot_longer(
    names_to = "drugs",
    values_to = "concentrations",
    cols = c(penicillin, streptomycin, neomycin)
  ) %>%
  ggplot(
    aes(
      x = drugs,
      y = bacteria,
      color = log(concentrations),
      shape = gram
    )
  ) +
  geom_point(size = 4) +
  scale_color_gradient(low = "Black", high = "Red") 
```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.3}
# WRITE YOUR CODE HERE
df_antibiotics %>%
  pivot_longer(
    names_to = "drugs",
    values_to = "concentrations",
    cols = c(penicillin, streptomycin, neomycin)
  ) %>%
  filter(drugs == "penicillin") %>%
  mutate(
    satisfactory = (concentrations < 0.1)
  ) %>%
  ggplot(
    aes(
      x = bacteria,
      y = concentrations,
      fill = satisfactory
    )
  ) +
  geom_col() +
  coord_flip() + 
  scale_y_log10() 
```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.4}
# WRITE YOUR CODE HERE
df_antibiotics %>%
  pivot_longer(
    names_to = "drugs",
    values_to = "concentrations",
    cols = c(penicillin, streptomycin, neomycin)
  ) %>% 
  filter(drugs == "penicillin") %>%
  ggplot(
    aes(
      x = concentrations,
      fill = gram,
      y = bacteria
    )
  ) + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    plot.title = element_text(hjust = 0.5)
    ) +
  geom_col(position = position_dodge()) +
  coord_flip() +
  scale_x_log10() +
  labs(title = "Penicillin Tests")

df_antibiotics %>%
  pivot_longer(
    names_to = "drugs",
    values_to = "concentrations",
    cols = c(penicillin, streptomycin, neomycin)
  ) %>% 
  filter(drugs == "streptomycin") %>%
  ggplot(
    aes(
      x = concentrations,
      fill = gram,
      y = bacteria
    )
  ) + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  ) +
  geom_col(position = position_dodge()) +
  coord_flip() +
  scale_x_log10() + 
  labs(title = "Streptomycin Tests")

df_antibiotics %>%
  pivot_longer(
    names_to = "drugs",
    values_to = "concentrations",
    cols = c(penicillin, streptomycin, neomycin)
  ) %>% 
  filter(drugs == "neomycin") %>%
  ggplot(
    aes(
      x = concentrations,
      fill = gram,
      y = bacteria
    )
  ) + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  ) +
  geom_col(position = position_dodge()) +
  coord_flip() +
  scale_x_log10() +
  labs(title = "Neomycin Tests")
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.5}
# WRITE YOUR CODE HERE
df_antibiotics %>%
  pivot_longer(
    names_to = "drugs",
    values_to = "concentrations",
    cols = c(penicillin, streptomycin, neomycin)
   ) %>%
  ggplot(
    aes(
      x = bacteria,
      y = concentrations,
      fill = drugs
    )
  ) +
  geom_col(position = position_dodge()) +
  coord_flip() +
  scale_y_log10()

```

### **q2** Assess your visuals

There are **two questions** below; use your five visuals to help answer both Guiding Questions.
Note that you must also identify which of your five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in answering the questions below.
You may need to revise one or more of your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most effective at helping answer both guiding questions.
**Use this as an opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against bacteria of different genera and Gram stain?

*Observations* - What is your response to the question above?
- Penicillin is more effective to bacteria with positive gram strain.
- Streptomycin is more effective to baterica with positive gram strain.
- Neomycin is more effective to bacteria with negative gram strain, although some positive strains are also highly effective.

-   Which of your visuals above (1 through 5) is **most effective** at helping to answer this question?

-   I mainly used the first plot, which was a scatterplot to visualize how the gram strain was related to the concentrations for all three antibiotics.

-   Why?

-   Since we want to compare all three antibiotics, it would make sense to look at visuals that explore all three at a time.
    My first visual had an x axis of logarithmic concentration, which allowed me to determine whether the MIC was satisfactory or not.
    I did create additional visuals to see if I could narrow down how each drug's concentration was related to the gram stain, but I found that it was more difficult to relate gram stain against the MIC concentrations.

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*, and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis* [2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*?

*Observations* - What is your response to the question above?
- From the data, I found that most bacteria classified as streptococcus were mainly gram negative.
Furthermore, most of these bacteria are susceptible to penicillin, but are resistant to the other antibiotics.
Looking at the bacteria pneumonia, we can see that pneumonia inhibits similar resistances to streptomycin and neomycin.

-   Which of your visuals above (1 through 5) is **most effective** at helping to answer this question?

-   For this graph, I found that visuals 1 and 3 were most helpful in identifying trends.

-   Why?

-   The overall scatterplot in visual 1 is useful in identifying how Streptococcus bacteria are similar to each other.
    From that graph, I was able to point out the gram stain similarities as well as antibiotic resistances.
    Visual 3 is more specific (only penicillin), which is helpful in verifying that penicillin is only useful against certain bacteria strains.

# References

<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide antibacterial range and rarely sensitizing.
Scope.
1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009) [link](https://www.americanscientist.org/article/thats-funny)
