---
title: "Massachusetts Highway Stops"
author: "Lucien Lin"
date: 4/28/2025
output:
  github_document:
    toc: true
---

*Purpose*: In this last challenge we'll focus on using logistic regression to study a large, complicated dataset. Interpreting the results of a model can be challenging---both in terms of the statistics and the real-world reasoning---so we'll get some practice in this challenge.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

*Background*: We'll study data from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/), specifically their dataset on Massachusetts State Patrol police stops.

```{r setup}
library(tidyverse)
library(broom)
```

# Setup

<!-- -------------------------------------------------- -->

### **q1** Go to the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/) page and download the Massachusetts State Police records in `Rds` format. Move the data to your `data` folder and match the `filename` to load the data.

*Note*: An `Rds` file is an R-specific file format. The function `readRDS` will read these files.

```{r q1-task}
## TODO: Download the data, move to your data folder, and load it
filename <- "data/yg821jf8611_ma_statewide_2020_04_01.rds"
df_data <- readRDS(filename)
```

# EDA

<!-- -------------------------------------------------- -->

### **q2** Do your "first checks" on the dataset. What are the basic facts about this dataset?

```{r}
df_data %>%
  summary()

df_data
```

**Observations**:

-   What are the basic facts about this dataset?
-   df_data contains 3416239 rows and 24 columns
-   the dataset contains information on the traffic stopped person. This information include race and sex. Information on when the stop occured is also given through the county and location(city) There is also information regarding the outcome of the traffic stop alongside additional details about whether contraband was found or not

Note that we have both a `subject_race` and `race_Raw` column. There are a few possibilities as to what `race_Raw` represents:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

Let's try to distinguish between these two possibilities.

### **q3** Check the set of factor levels for `subject_race` and `raw_Race`. What do you note about overlap / difference between the two sets?

```{r q3-task}
## TODO: Determine the factor levels for subject_race and raw_Race
unique(df_data$subject_race)

unique(df_data$raw_Race)
```

**Observations**:

-   What are the unique values for `subject_race`?
    -   white, hispanic, black, asian/pacific islander other, NA, unknown
-   What are the unique values for `raw_Race`?
    -   White, Hispanic, Black, Asian or Pacific Islander, Middle Eastern or East Indian (South Asian), American Indian or Alaskan Native, NA, None
-   What is the overlap between the two sets?
    -   white, hispanic, black, asian/pacific islander other, NA, unknown
-   What is the difference between the two sets?
    -   subject_race has less race classifications, where races such as Middle Eastern or American Indian aren't values for subject_race

### **q4** Check whether `subject_race` and `raw_Race` match for a large fraction of cases. Which of the two hypotheses above is most likely, based on your results?

*Note*: Just to be clear, I'm *not* asking you to do a *statistical* hypothesis test.

```{r q4-task}
## TODO: Devise your own way to test the hypothesis posed above.

df_data %>%
  mutate(
    matches = tolower(subject_race) == tolower(raw_Race)
  ) %>%
  summarize(
    matches = mean(matches, na.rm = TRUE)
  )
```

**Observations**

Between the two hypotheses:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

which is most plausible, based on your results?

-   With a match rate of 94%, I think that the race_Raw could be the unprocessed version of the subject_Race. This is because the percentage is really high, and that the 6% could be due to the additional values in race_Raw not matching with subject_Race.

## Vis

<!-- ------------------------- -->

### **q5** Compare the *arrest rate*---the fraction of total cases in which the subject was arrested---across different factors. Create as many visuals (or tables) as you need, but make sure to check the trends across all of the `subject` variables. Answer the questions under *observations* below.

(Note: Create as many chunks and visuals as you need)

```{r}
df_data %>%
  group_by(subject_race) %>%
  summarize(
    arrest_rate = mean(arrest_made, na.rm = TRUE),
  ) %>%
  ggplot(
    aes(
      x = subject_race, 
      y = arrest_rate
    )
  ) + 
  geom_col()

df_data %>%
  group_by(subject_age) %>%
  summarize(
    arrest_rate = mean(arrest_made, na.rm = TRUE)
  ) %>%
  ggplot(
    aes(
      x = subject_age, 
      y = arrest_rate
    )
  ) + 
  geom_col()

df_data %>%
  group_by(subject_sex) %>%
  summarize(
    arrest_rate = mean(arrest_made, na.rm = TRUE)
  ) %>%
  ggplot(
    aes(
      subject_sex, 
      arrest_rate
    )
  ) + 
  geom_col()
```

**Observations**:

-   How does `arrest_rate` tend to vary with `subject_age`?
    -   The arrest rate was high for folks under 25, with a large spike occuring at around the 17-18 age.
-   How does `arrest_rate` tend to vary with `subject_sex`?
    -   Males tended to have a higher arrest rate compared to females.
-   How does `arrest_rate` tend to vary with `subject_race`?
    -   Hispanics tend to have a higher arrest rates than others. We can see that compared to white, hispanics had almost triple the arrest rates.

# Modeling

<!-- -------------------------------------------------- -->

We're going to use a model to study the relationship between `subject` factors and arrest rate, but first we need to understand a bit more about *dummy variables*

### **q6** Run the following code and interpret the regression coefficients. Answer the the questions under *observations* below.

```{r q6-task}
## NOTE: No need to edit; inspect the estimated model terms.
fit_q6 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )

fit_q6 %>% tidy()
```

**Observations**:

-   Which `subject_race` levels are included in fitting the model?
    -   Hispanic, white, black
-   Which `subject_race` levels have terms in the model?
    -   Hispanic, white

You should find that each factor in the model has a level *missing* in its set of terms. This is because R represents factors against a *reference level*: The model treats one factor level as "default", and each factor model term represents a change from that "default" behavior. For instance, the model above treats `subject_sex==male` as the reference level, so the `subject_sexfemale` term represents the *change in probability* of arrest due to a person being female (rather than male).

The this reference level approach to coding factors is necessary for [technical reasons](https://www.andrew.cmu.edu/user/achoulde/94842/lectures/lecture10/lecture10-94842.html#why-is-one-of-the-levels-missing-in-the-regression), but it complicates interpreting the model results. For instance; if we want to compare two levels, neither of which are the reference level, we have to consider the difference in their model coefficients. But if we want to compare all levels against one "baseline" level, then we can relevel the data to facilitate this comparison.

By default `glm` uses the first factor level present as the reference level. Therefore we can use `mutate(factor = fct_relevel(factor, "desired_level"))` to set our `"desired_level"` as the reference factor.

### **q7** Re-fit the logistic regression from q6 setting `"white"` as the reference level for `subject_race`. Interpret the the model terms and answer the questions below.

```{r q7-task}
## TODO: Re-fit the logistic regression, but set "white" as the reference
## level for subject_race

fit_q7 <- 
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ) %>%
      mutate(subject_race = fct_relevel(subject_race, "white")),
    family = "binomial"
  )

fit_q7 %>% tidy()
```

**Observations**:

-   Which `subject_race` level has the highest probability of being arrested, according to this model? Which has the lowest probability?
    -   According to this model, Hispanics have the highest probability of being arrested while whites had the lowest probability of being arrested since it was used as reference.
-   What could explain this difference in probabilities of arrest across race? List **multiple** possibilities.
    -   Socio-economic status
    -   Racial prejudice
    -   Types of offense
-   Look at the set of variables in the dataset; do any of the columns relate to a potential explanation you listed?
    -   There is a column that explains reason_for_stop which supports the explanation for types of offense.

One way we can explain differential arrest rates is to include some measure indicating the presence of an arrestable offense. We'll do this in a particular way in the next task.

### **q8** Re-fit the model using a factor indicating the presence of contraband in the subject's vehicle. Answer the questions under *observations* below.

```{r q8-task}
## TODO: Repeat the modeling above, but control for whether contraband was found
## during the police stop

fit_q8 <- 
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex + contraband_found,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ), 
    family = "binomial"
  )

fit_q8 %>% tidy()
```

**Observations**:

-   How does controlling for found contraband affect the `subject_race` terms in the model?
    -   Contraband found lowers the subject_race terms.
-   What does the *finding of contraband* tell us about the stop? What does it *not* tell us about the stop?
    -   When you find contraband, there is a valid reason for arrest, which can lead to higher rates of actual arrest.
    -   Contraband found doesn't provide information on how that contraband was found, whether through visible inspection, or a vehicle search. An officer is the one who can decide whether a search is needed, which can present bias.

### **q9** Go deeper: Pose at least one more question about the data and fit at least one more model in support of answering that question.

```{r}
fit_q9 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex + contraband_found + vehicle_type,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ), 
    family = "binomial"
  )
fit_q9 %>% tidy()
```

**Observations**:

-   Question: Does the vehicle type have an effect on arrest rate?
-   Observations:
    -   I notice that motorcycle stops have a really high correlation with arrest rates.
    -   Passenger vehicles also have a positive value for the arrest_rates correlation. However, I would like to know what the percentage of passenger vehicles is in this dataset, since passenger cars are probably the most popular vehicle type. It is possible that all races drive some sort of passenger cars, which may mean that the subject races are more indicative of arrests, though this table says otherwise (due to the mentioned possibility that most of the arrests were made for folks driving passenger cars)

## Further Reading

<!-- -------------------------------------------------- -->

-   Stanford Open Policing Project [findings](https://openpolicing.stanford.edu/findings/).
